---
title: "New York City flights"
description: |
  An analysis of the `nycflights13` datasets. Mandatory project report in Tools for Analytics (R part).
author:
  - name: [First author name]
    #url: https://example.com/norajones
    affiliation: cand.merc (OSCM)
    #affiliation_url: https://example.com/spacelysprokets
  - name: [Second author name]
    affiliation: cand.merc (?)
  - name: [Third author name]
    affiliation: cand.merc (?)
date: "`r Sys.Date()`"
#repository_url: [A url to your source code(dropbox, onedrive, github etc)]
#preview: https://rstudio.github.io/distill/images/javascript-d3-preview.png
creative_commons: CC BY-NC
output:
  distill::distill_article:
    toc: true
    toc_depth: 3
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

```{r, include=FALSE}
if (interactive()) setwd(dirname(rstudioapi::getActiveDocumentContext()$path)) # set working dir to current file location
knitr::opts_chunk$set(
  cache = TRUE, autodep = TRUE,  # use this option so compile the document faster (you may comment it out at the final report)
  echo = TRUE, 
  layout="l-page", fig.width = 12
  )
# use xaringan::inf_mr() for instant preview
```

<!-- Some css for the comments (delete) -->
```{css, echo=FALSE}
.comment {
  color: #F1948A;
  border-color: black;
  border-style: dashed;
  border-width: thin;
  padding: 10px;
}
```

## Introduction

We consider the datasets available from the package `nycflights13` that contains information about every flight that departed from New York City in 2013. Let us have a look at the datasets. First, we load the packages need for this report:

```{r, cache=FALSE}
library(tidyverse)
library(nycflights13)
library(skimr)
library(knitr)
library(kableExtra)
library(directlabels)
library(patchwork)
library(rmarkdown)
library(lubridate)
```

The datasets in the `nycflights13` package are:

```{r, fig.align='center', echo=FALSE}
res <- data(package = "nycflights13", verbose = T)
res$results %>% 
  as_tibble() %>% 
  select(Dataset = Item, Description = Title) %>% 
  kable() %>% 
  kable_styling(position = "center")
```

Let us try to do some descriptive analytics on the different datasets.



## Flights

In this section we will focus on the `flights` data set, which lists all domestic flights out of the New York area in 2013. We run `skim` to get an overview:

```{r}
skim(flights)
```

The variables in this dataset are:

* `year, month, day` Date of departure
* `dep_time,arr_time` Actual departure and arrival times.
* `sched_dep_time, sched_arr_time` Scheduled departure and arrival times.
* `dep_delay, arr_delay` delays in minutes
* `hour, minute` Time of scheduled departure
* `carrier` carrier abbreviation
* `tailnum` Tail number of plane.
* `flight` flight number.
* `origin, dest` Origin and Destination
* `air_time` Time spent in air.
* `distance` Distance flown.
* `time_hour` scheduled date and hour of flight.

For further details about the dataset see `?flights` or the [online documentation](https://www.rdocumentation.org/packages/nycflights13/versions/1.0.1/topics/flights).

The skim output indicates that some flights are canceled. We remove these observations from the dataset:

```{r}
dat <- flights %>%
  filter(!is.na(dep_time))
```

### Joining datasets

Let us first try to do some [mutating joins](https://bss-osca.github.io/tfa/sec-transform.html#mutating-joins) and combine variables from multiple tables. In `flights` we have flight information with an abbreviation for carrier (`carrier`), and in `airlines` we have a mapping between abbreviations and full names (`name`). You can use a join to add the carrier names to the flight data:

```{r, warning = FALSE}
dat <- dat %>% 
  left_join(airlines) %>% 
  rename(carrier_name = name) %>% 
  print()
```

Note we here join by the column `carrier` represented in both data frames. That is, the default argument `by = c("carrier" = "carrier")` is used. If we want the full name of origin airport, we need to specify which one we want to join to since each flight has an origin and destination `airport`. Afterwards we do the same for the destination airport. 

```{r}
dat <- dat %>% 
  left_join(airports %>% select(faa, name), 
            by = c("origin" = "faa")) %>% 
  rename(origin_name = name) %>% 
  left_join(airports %>% select(faa, name), 
            by = c("dest" = "faa")) %>% 
  rename(dest_name = name) %>% 
  select(month, carrier_name, origin_name, dest_name, sched_dep_time, dep_delay, arr_delay, distance, tailnum) %>% 
  print()
```

We now have the flights data we need stored in the data frame `dat`. Let us try to answer some questions.


### How many flights leave each New York airport for each carrier? 

We first calculate a summary table:

```{r}
dat %>% 
  count(origin_name, carrier_name, sort = TRUE) %>% 
  paged_table()
```

Let us visualize the numbers. First we facet by airport and use `geom_bar`:

```{r, fig.asp=0.75}
dat %>% 
  ggplot(aes(carrier_name)) +
  geom_bar() + 
  facet_grid(rows = vars(origin_name)) + 
  labs(
    title = "Number of flights",
    x = "Carrier",
    y = "Flights"
  ) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

We can also compare the two categorical variables by using `geom_count`:

```{r, eval=FALSE}
# SOLVED
dat %>%
  ggplot(aes(carrier_name, origin_name)) + 
  geom_count() + 
  labs(
    title = "Number of flights",
    y = "Carrier",
    x = "Departure airport",
    size = "Flights"
   ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Finally, we can use a heatmap by using `geom_tile`. In this case, `geom_tile` doesn't offer a way to calculate counts on it's own, so we use the function `count` in our pipe:

```{r}
dat %>%
  count(origin_name, carrier_name) %>%
  ggplot(aes(origin_name, carrier_name, fill = n)) + 
  geom_tile() + 
  labs(
    title = "Number of flights",
    y = "Carrier",
    x = "Departure airport",
    fill = "Flights"
  ) 
```



### How many carrier flights per month?

Summaries are: 

```{r}
# SOLVED
dat %>%
  count(month, carrier_name) %>%
  paged_table()
```
We will try to visualize the numbers using a line plot with carrier as color aesthetic:

<aside>
We here will use the package *directlabels* to add the labels directly beside the lines. See [here]( http://directlabels.r-forge.r-project.org/docs/lineplot/plots/lars.html) for info.   
</aside>

```{r}
# SOLVED (Don't like the labels, look into cleaning that up)
dat %>%
  count(month, carrier_name) %>%
  ggplot(mapping = aes(x = month, y = n, color = carrier_name)) +
  geom_line() +
  geom_point() +
  geom_dl(aes(label = carrier_name), method = list(dl.trans(x = x + .3), "last.bumpup")) + 
  scale_x_continuous(breaks = 1:12, limits = c(1,17)) + 
  labs(
    title = "Number of flights",
    y = "Flights",
    x = "Month"
  ) + 
  theme(legend.position = "none") 
```

### Which carriers/airlines have the worst delays?

Note that the delays columns are in minutes. We first convert delays to hours:

```{r}
dat <- dat %>% 
  mutate(across(contains("delay"), ~ .x / 60))
```

Next, we answer the question by looking at different measures.

#### Average delay

Let us first have a look at the average departure delay by airline. The `dplyr` package has two functions that make it easy to do that: the `group_by` and the `summarize` functions. We use the two together and groups the rows of the dataset together based on the `carrier` and then uses `summarise` and the `mean` function to calculate the average delay:

```{r}
dat %>%
  group_by(carrier_name) %>%
  summarise(ave_delay = mean(dep_delay, na.rm = TRUE)) %>%
  arrange(desc(ave_delay)) %>%
  paged_table()
```

Note the `mean` function has a `na.rm` argument which ignores the missing values otherwise the average delays could not be calculated. We can visualize our summary (a continuous-categorical comparison) by piping the table into a column plot:

```{r}
dat %>% 
  group_by(carrier_name) %>% 
  summarise(ave_delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(carrier_name, ave_delay)) + 
  geom_col()
```
To get a better visualization we reorder the categorical x-axis by average delay, use the full names of the airlines (which are rotated) and add some informative labels:

```{r, echo=TRUE}
# SOLVED
dat %>% 
  group_by(carrier_name) %>% 
  summarise(ave_delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(carrier_name, -ave_delay),y = ave_delay)) + 
  geom_col() +
  labs(title = "Average delay of airline carriers", 
       x = "Airline carrier", 
       y = "Average delay in hours") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

To conclude, Frontier (F9) and Express Jet (EV) have the highest average delay. However, using `mean` to summarize a value can be dangerous, because it's sensitive to outliers!





#### Variation

We should *always* ask about the variation in the variables in our data sets, but it's especially important to do so if we're going to use averages to summarize them.

First let us calculate the standard deviation for each carrier:

```{r}
# SOLVED
dat %>% 
  group_by(carrier_name) %>% 
  summarise(std_delay = sd(dep_delay, na.rm = TRUE)) %>%
  arrange(desc(std_delay)) %>%
  paged_table()
```

What is the distribution of departure delays by airline? Visualized as a density distribution using carrier as fill aesthetic:

```{r warning=FALSE}
# SOLVED
dat %>%
  ggplot(aes(dep_delay, fill = carrier_name)) + 
  geom_density(alpha = 0.5) + 
  labs(
    title = "Distribution of departure delays",
    x = "Delay (hours)",
    y = "Density",
    fill = "Carrier"
  ) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

We can see that there is a small number of HUGE outliers which makes using `mean` possibly very misleading. 


Lets us try to make a plot of the empirical cumulative distributions for each carrier using carrier as color aesthetic and a zoom of at most 3 hours delay:

<aside>
Have a look [here](http://rstudio-pubs-static.s3.amazonaws.com/209392_437ec4da7fa2432d831320f3591e7491.html) to see how to zoom.
</aside>

```{r}
# SOLVED (should we limit c(-3, 3) instead?)
dat %>%
  ggplot() + 
  stat_ecdf(aes(x = dep_delay, color = carrier_name), alpha = 0.75) +
  coord_cartesian(xlim = c(-0.5, 3)) +
  labs(title = "Empirical cumulative distribution of departure delays", 
       x = "Delays (hours)", 
       y = "Cumulative distribution")
```
Note, the higher upper-left the distribution is, the better. That is, a carrier dominates other carriers if the line is above the other carriers. Comparing this to the standard deviations, we see that the standard deviations is not a good measure for delays.

Variation in data like these where the outliers are very sparse is hard to visualize using density plots. We may also use a boxplot:

```{r}
# SOLVED
dat %>% 
  ggplot(aes(x = carrier_name, y=dep_delay)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

We can see that most flights have a median around zero. However, some carriers have larger delays compared to others. Is the variation in departure delay different given departure airport? We use departure airport as color aesthetic:

```{r}
# WIP (finish commenting on results below)
dat %>% 
  ggplot(aes(x = carrier_name, y=dep_delay, color = origin_name)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

[Comment on the results here]
There doesn't seem to be a significant difference between the 3 airports. One possible observation is that John F Kennedy Intl has more extreme outliers compared to the other airports.




#### Median

The boxplot shows median values in the center. What would happen if we used `median` instead of average delay time and make a column plot? 

```{r}
# SOLVED (check comment on results below) 
dat %>% 
  group_by(carrier_name) %>% 
  summarise(median_delay = median(dep_delay)) %>%
  ggplot(aes(carrier_name, median_delay)) + 
  geom_col() + 
  labs(
    title = "Median departure delay for each carrier",
    x = "Carrier",
    y = "Median delay (hours)"
  ) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
That tells a bit of a different story!  [Comment on the results here] All previous results based on average showed rather high delays for all cariers (most between 0.1 and 0.3 hours). Looking at the median however shows that the few extreme delays heavily skewed the average delay. We now see most that carriers even have a median departure time that's slightly ahead of schedule, while a few carriers have a very minor actual median delay time (less than 0.025 hours, or 1.5 minutes).



#### Delays of more than an hour

How many flights were **really** delayed and how does that break down by airline carrier?  Being delayed more than an hour really sucks, so let's use that as our cutoff:

```{r}
# SOLVED
dat %>% 
  filter(dep_delay >= 1) %>% 
  count()
```

That's a lot of flights!  We can use the `dplyr` function named `count` to give us a summary of the number of rows of a that correspond to each carrier:

```{r}
# SOLVED
dat %>% 
  filter(dep_delay >= 1) %>% 
  group_by(carrier_name) %>% 
  count(carrier_name, sort = TRUE) %>% 
  paged_table()
```

Note that `count` has created a column named `n` which contains the counts and we ask it to sort that column for us.

We can visualize it with a column plot (note we don't need to reorder because `count` has done that for us):

```{r}
# SOLVED (I do need to reorder though... Something wrong here?)
# I'm also a bit lost with what to comment on the results below...
dat %>% 
  filter(dep_delay >= 1) %>% 
  group_by(carrier_name) %>% 
  count(carrier_name, sort = TRUE) %>%
  ggplot(aes(x = reorder(carrier_name, -n),y = n)) +
  geom_col() +
  labs(title = "Number of delays over 1 hour for each airline carrier", 
       x = "Carrier", 
       y = "Number of delays over 1 hour") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

[Comment on the results here] ExpressJet Airlines has a serious problem with long delays. One has be cautious when drawing conclusions here though. We don't know the total number of flights for each carrier, so it might be that ExpressJet just performs way more flights and therefore has more delays.



### What is the relationship between departure delay and arrival delay?

We plot the delays against each other as points. 

```{r}
# SOLVED
dat %>% 
  ggplot(aes(dep_delay, arr_delay)) +
  geom_point(alpha = 0.1) +
  labs(title = "Arrival delays vs. Departure delays", 
       x = "Departure delays (Hours)", 
       y = "Arrival delays (Hours")
```

The large mass of points near (0, 0) can cause some confusion since it is hard to tell the true number of points that are plotted. This is the result of a phenomenon called overplotting. As one may guess, this corresponds to points being plotted on top of each other over and over again. When overplotting occurs, it is difficult to know the number of points being plotted. We adjust the transparency of the points by setting `alpha = 0.1`

[Comment on the results here] As would be logical, it seems departure delays heavily correlate with arrival delays. For the relatively low departure delays (<8 hours) it does seem, however, that the corresponding arrival delays tend to be larger than the departure delays. One possible explanation is that most of these delays could be weather-related, and therefore a longer flight time results in 1) the aircraft being late from a previous flight (departure delay) and 2) the aircraft being even later on arrival due to the longer flight time. This cannot be observed for the very large delays (over 12 hours). That might be due to the low amount of data points there, or because the delays have other reasons than the weather and flight time.


### Are flight delays worse at different New York airports? 

If you're flying out of New York you might want to know which airport has the worst delays on average. We first calculate median and average delays:

```{r}
# SOLVED
dat %>% 
  group_by(origin_name) %>% 
  summarise(med_delay = median(dep_delay), ave_delay = mean(dep_delay))
```

As we can see La Guardia seems to have the smallest delays. However, the difference is small. Lets us try to make a plot of the empirical cumulative distributions for each airport using airport as color aesthetic and a zoom of at most 2 hours:

```{r}
# WIP
dat %>%
  ggplot(aes(x = dep_delay, color = origin_name)) + 
  stat_ecdf(alpha = 0.7) +
  coord_cartesian(xlim = c(-0.5, 2)) +
  labs(title = "Empirical cumulative distribution of departure delays", 
       x = "Delays (hours)", 
       y = "Cumulative distribution")
```

The median values can be found at y = 0.5. Note that La Guardia is above the other lines indicating that it has the smallest delays no matter what fractile we consider. Another way to visualize this covariation in a categorical (airport) and a continuous (delay) variable is with a boxplot. We use a little scaling to get a better picture of the average delay and zoom so the y variable is between at most half an hour.

```{r}
# WIP - what does he mean "We use a little scaling to get a better picture of the average delay". Something with scale_y_continuous I suppose?
dat %>% 
  ggplot(aes(x = origin_name, y = dep_delay)) +
  geom_boxplot() +
  #scale_y_continuous() +
  coord_cartesian(ylim = c(-0.5, 0.5))
```


### Are carrier flight delays different at New York airports? 

We first calculate median and average delays:

```{r}
# SOLVED (not too sure about the grouping here?)
dat %>% 
  group_by(origin_name, carrier_name) %>% 
  summarise(med_delay = median(dep_delay), ave_delay = mean(dep_delay)) %>% 
  paged_table()
```

There are some differences. Let us try to do a heat map of the average delays:

```{r}
# SOLVED
dat %>%
  group_by(origin_name, carrier_name) %>%
  summarise(ave_delay = mean(dep_delay)) %>%
  ggplot(aes(origin_name, carrier_name, fill = ave_delay)) +
  geom_tile() +
  scale_fill_continuous(low = "#31a354", high = "#e5f5e0") + 
  labs(
    title = "Average departure delays",
    x = "Departure airport",
    y = "Carrier",
    fill = "Ave. delay (min)"
  ) 
```
For each carrier this give a good insight into the differences at each airport. Another way to visualize the covariation is with a boxplot. We use a little scaling to get a better picture of the delay and zoom so the delay is at most half an hour.

```{r}
# WIP - Again, what does he mean "We use a little scaling to get a better picture of the  delay". Something with scale_y_continuous I suppose?

# Also, I tried different things (origin on color etc.), but they were all really messy. So it's this facet_grid for now. Maybe something else fits better?
dat %>% 
  ggplot(aes(x = carrier_name, y = dep_delay)) +
  geom_boxplot() +
  #scale_y_continuous() +
  facet_grid(rows = vars(origin_name)) +
  coord_cartesian(ylim = c(-0.5, 0.5)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

We may also try to plot the empirical cumulative distributions for each carrier (facet) using airport as color aesthetic and a zoom of the delay is at most 1 hour:

```{r}
# SOLVED (is this what he wants? Looks a bit messy?)
dat %>% 
  ggplot(aes(x = dep_delay, color = origin_name)) +
  stat_ecdf(alpha = 0.7) +
  coord_cartesian(xlim = c(-0.5, 1)) +
  labs(title = "Empirical cumulative distribution of departure delays", 
       x = "Delays (hours)", 
       y = "Cumulative distribution",
       color = "Departure airport") +
  facet_wrap(facets = vars(carrier_name)) +
  theme(legend.position = "bottom")
```



### Does departure time affect flight delays? 

First, note that the `sched_dep_time` is a number in the format HHMM. We convert it into a hour:minutes data type and afterwards to hours since midnight:

```{r}
dat <- dat %>% 
  mutate(sched_dep_time = hm(str_replace(sched_dep_time, "^(.*)(..)$", "\\1:\\2"))) %>% 
  mutate(sched_dep_time = as.numeric(sched_dep_time)/60/60)
```


To explore covariation in two continuous (quantitative) variables, we can use a scatter plot:

```{r}
# SOLVED
dat %>% 
  ggplot(aes(sched_dep_time, dep_delay)) +
  geom_point(alpha = 0.1) +
  labs(title = "Departure delays vs. Scheduled departure time", 
       x = "Schedule departure time (Time in 24h)", 
       y = "Delays (Hours)")
```

[Comment on the results here] First, we can observe several "towers". There are most likely standard set departure times so they have more flights and therefore more delays. Second, it would seem flights scheduled after 15:00 experience more delays than morning flights.



### Does travel distance affect departure and arrival delay?

We use the patchwork package to plot distance against the two delays. Moreover we also add a smoothed line using `geom_smooth`:

```{r}
# SOLVED - smooth line seems useless. Not sure what to comment tbh.
p1 <- dat %>% 
  ggplot(aes(x=distance, y= dep_delay)) + 
  geom_point() +
  geom_smooth() +
  labs(title = "Departure delays vs. Distance",
       x = "Distance (km)", 
       y = "Departure delay (Hours)")

p2 <- dat %>% 
  ggplot(aes(x=distance, y= arr_delay)) + 
  geom_point() +
  geom_smooth() +
  labs(title = "Arrival delays vs. Distance",
       x = "Distance (km)", 
       y = "Arrival delay (Hours)")

p1 + p2
```

[Comment on the results here] Results seems very similar for departure and arrival times.







## Planes

Let us do a mutation join so we have a bit more information about each airplane:

```{r}
dat <- dat %>% 
  left_join(planes %>% 
              select(tailnum, plane_manufacturer = manufacturer, plane_model = model))
```

### What is the monthly usage of all the aircrafts? 

This could be useful for some kind of maintenance activity that needs to be done after x number of trips. The summary table is (based on `tailnum`):

```{r}
# SOLVED
dat %>% 
  group_by(tailnum) %>% 
  count(tailnum, sort = TRUE) %>% 
  paged_table()
```

As an example, consider the plane N355NB:

```{r}
dat1 <- dat %>% 
  filter(tailnum=="N355NB") 
```

The specifications are:

```{r}
filter(planes, tailnum=="N355NB")
```

We see that it is an Airbus 319 with 145 seats. The plane flew `r nrow(dat1)` flights in 2013 with a total distance of `r sum(dat1$distance)`.   
Let us have a look at the destinations:

```{r}
# SOLVED
dat1 %>% 
  count(dest_name) %>% 
  ggplot(aes(x = reorder(dest_name, -n), n)) + 
  geom_col() +
  labs(title = "Number of flights per destination for N355NB", 
       x = "Destination airport", 
       y = "Number of flights") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```


## Weather

In this section we will focus on the `weather` data set, which lists hourly metreological data for LGA, JFK and EWR. We run `skim` to get an overview:

```{r}
skim(weather)
```

For further details see `View(weather)` or read the associated help file by running `?weather` to bring up the help file.

Observe that there is a variable called temp of hourly temperature recordings in Fahrenheit at weather stations near all three major airports in New York City: Newark (origin code EWR), John F. Kennedy International (JFK), and LaGuardia (LGA). Let us transform the temperature to celsius:

```{r}
dat_w <- weather %>% 
  left_join(airports %>% select(faa, name), 
            by = c("origin" = "faa")) %>% 
  rename(origin_name = name) %>% 
  mutate(temp = (temp - 32) * (5/9) ) %>% 
  select(origin_name, time_hour, month, temp)
```

### How are the temperatures fluctuating over the year?

We start by plotting temperature over the year with airport/origin as color aesthetic. We also add a smoothing line:

```{r}
dat_w %>% 
  ggplot(aes(x = month, y = temp, color = origin_name)) +
  geom_point(alpha = 0.1) +
  geom_smooth(alpha = 0.1) +
  labs(title = "Temperature fluctuation at NYC airports throughout the year 2013",
       x = "Month (number)",
       y = "Temp (C째)")
```

Note that we have used the `alpha` aesthetic to make the lines more transparent. [Comment on the results here] JFK has a lower temperature throughout spring and summer. Otherwise the airports are pretty similar. I know JFK is close to the water, so the that might account for the slightly lower temperature.

### Are the temparatures different in the airports? 

Let us start by plotting the density for each airport:

```{r}
# SOLVED (but not sure if right tbh, his interpretation seems to talk about something else)
dat_w %>% 
  ggplot(aes(x = temp, fill = origin_name)) +
  geom_density(alpha = 0.4) +
  labs(title = "Temperature density for each NYC airport", 
       x = "Temperature (C째)", 
       y = "Density", 
       color = "Airport")
```

Note the mean temparature is more or less the same (vertical lines). There is a bit fluctuations on Newark compared to for instance JFK airport (lowest spread).

A closer look can be done by faceting by month:

```{r}
dat_w %>% 
  ggplot(aes(x = temp, fill = origin_name)) +
  geom_density(alpha = 0.4) +
  facet_wrap(vars(month)) +
  labs(title = "Temperature density for each NYC airport, per month", 
       x = "Temperature (C째)", 
       y = "Density", 
       color = "Airport")
```

Finally, let us consider a boxplot of temparature for each month:

```{r}
# SOLVED (still needs commenting below)
dat_w %>% 
  ggplot(aes(y = temp)) +
  geom_boxplot() +
  facet_wrap(vars(month)) +
  labs(title = "Temperature boxplots NYC airports per month", 
       y = "Temperature (C째)") +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank())
```

[Comment on the results here]

What does the dot at the bottom of the plot for May correspond to?  Explain what might have occurred in May to produce this point.



## Any insights on canceled flights?

The canceled flights are:

```{r}
dat_c <- flights %>%
  filter(is.na(dep_time))
```

Let us do some analysis.

<aside>
Add a few plots you think are important.
</aside>

## Other insights?

<aside>
Include further analysis you think are important.
</aside>





## Colophon 
<!-- Always keep this section -->

This report has been created inside [RStudio](http://www.rstudio.com/ide/) using [R Markdown](https://rmarkdown.rstudio.com/) and the [distill](https://rstudio.github.io/distill/) format. 

The report was built using:

```{r message = FALSE, warning = FALSE, echo = FALSE}
session <- devtools::session_info()
session$platform
```

Along with these packages:

```{r message = FALSE, warning = FALSE, echo = FALSE}
session$packages %>% 
  select(package, loadedversion, date, source) %>% 
  DT::datatable(rownames = FALSE,
                class = 'cell-border stripe',
                filter = list(position = 'top'),
                options = list(pageLength = 5, 
                           autoWidth = FALSE,
                           bInfo = FALSE,
                           paging = TRUE))
```








