---
title: "Exercises Module 7"
author:
  - name: Gillian Loones
output: 
  html_document:
    toc: true
    toc_depth: 2
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
# set default chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  autodep = TRUE
  )
```

<!-- Some css for the comments (delete) -->
```{css, echo=FALSE}
.comment {
  color: gold;
  border-style: dashed;
  border-width: thin;
  padding: 10px;
}
```


## Load packages:

```{r}
library(tidyverse)
library(tfa)
```


## Exercise 7.10.1 - Statistikbanken

You can use the API from Statistikbanken to download a lot of data sets. Let us consider airports in Denmark (data set with table id FLYV41):

```{r}
url <- "https://api.statbank.dk/v1/data/FLYV41/CSV?lang=en&LUFTHAVN=*&Tid=*&Transport=*"
```

**Question 1 - Use `cat(read_lines(url, n_max = 3), sep = "\n")` to have a look at the delimiter used.**

```{r}
cat(read_lines(url, n_max = 3), sep = "\n")
```


**Question 2 - Import the csv file.**

```{r}
airports <- read.csv2(url)

str(airports)
head(airports)
glimpse(airports)
```


**Question 3- Try to retrieve information and get an overview over the data by running:**

```{r}
library(jsonlite)
url <- "https://api.statbank.dk/v1/tableinfo/FLYV41?lang=en"
lst <- read_json(url, simplifyVector = T)
View(lst)
```

Note the data returned is in json format, so we use `read_json` to read the data into a list.


**Question 4 - Create a function `info(tab_id)` that returns a list with components `description`, `unit` and `variables` from the information for a data set with table id `tab_id`.**

```{r}
info <- function(tab_id) {
  # FLYV41 is table ID in previous question, so replace that part in the url with function input tab_id
  url <- str_c("https://api.statbank.dk/v1/tableinfo/", tab_id, "?lang=en")
  lst <- read_json(url, simplifyVector = T)
  # For variables, returning only the 1st 2 columns because only they contain relevant information (variable names)
  return(list(description = lst$description, unit = lst$unit, variables = lst$variables[,1:2]))
}

info("FLYV41")
```


**Question 5 - Information about all the data sets can be retrieved using:**

```{r}
url <- "https://api.statbank.dk/v1/tables?lang=en"
lst <- jsonlite::read_json(url, simplifyVector = T)
View(lst)
```

Have a look at FLYV41.


Given the information about variables in a data set we can construct the url to retrieve the data in csv format:

```{r}
tab_id <- "FLYV41"
url <- str_c("https://api.statbank.dk/v1/tableinfo/", tab_id, "?lang=en")
lst <- read_json(url, simplifyVector = T) 
col_id <- c(1,3)  # column ids in lst$variables$id
cols <- lst$variables$id[col_id]
url <- str_c("https://api.statbank.dk/v1/data/", tab_id, "/CSV?lang=en&", str_c(cols, collapse = "=*&"), "=*") %>% 
   URLencode()
url
```

**Question 6 - Create a function `get_data(tab_id, col_id)` that retrieve a data set.**

```{r}
get_data <- function(tab_id, col_id = NULL) {
  url <- str_c("https://api.statbank.dk/v1/tableinfo/", tab_id, "?lang=en")
  lst <- read_json(url, simplifyVector = T)
  cols <- lst$variables$id
  if (!is.null(col_id)) cols <- cols[col_id]
  
  url <- str_c("https://api.statbank.dk/v1/data/", tab_id, "/CSV?lang=en&",
               str_c(cols, collapse = "=*&"), 
               "=*") %>% URLencode()
  
  dat <- read.csv2(url)
  return(dat)
}

get_data("FLYV41", 3)
get_data("FLYV41", 1:2)
get_data("FLYV41")
```



**Question 7 - Use the function get_data to retrieve data for `tab_id = "FOLK1A"` and `col_id = c(2, 3, 5)` and save it as a csv file with a comma as delimiter.**

```{r}
dat1 <- get_data("FOLK1A", c(2, 3, 5))

dat1 %>% write.csv(file = "folk1a.csv")
```


**Question 8 - Save the data in an Excel file and a Google sheet.**

```{r}
library(openxlsx)
write.xlsx(dat1, file = "folk1a.xlsx")

library(googlesheets4)
gs4_auth() # doesn't work and can't be bothered to figure out why. Not important.
gs <- gs4_create("folk1a")
write_sheet(dat1, ss = gs)
gs4_browse(gs)  # have a look at the file in a browser
```



## Exercise 7.10.2 - Tuples in OPL

In the algebraic modeling language OPL (Optimization Programming Language) used by IBM ILOG CPLEX Optimization Studio, you can define tuples to contain various information. For example consider tuples defined as:

```
tuple nurse {
  string name;
  int experience;   // higest best
}

tuple shift {
   string departmentName;
   string day;
   int startTime;
   int endTime;
}
```

A `nurse` tuple is then defined as `<"Anne", 11>` and a `shift` tuple as `<“Consultation”, “Monday” 12, 18>`.

A set of tuples can be defined using:

```
{nurse} nurses = ...;
{shift} shifts = ...;
```

where the `...` operator means that the sets are read from a data text file:

```
nurses = {
   <"Anne", 11>, 
   <"Bethanie", 4>, 
   <"Betsy", 2>
};

shifts = {
   <"Emergency", "Monday", 2, 8>,
   <"Emergency", Monday 8 12 4 7>, 
   <"Emergency", "Monday" 12 18 2 5> 
};
```

You can now use the sets to define decision variables Xns equal one if nurse n is assigned to shift s.

In this exercise we will try to generate the data text file given tibbles with data.

**Question 1 - Try to generate a text file named test.dat using function write_lines with content**

```
nurses = {
   <"Anne", 11>
};
```

```{r}
write_lines(x = "nurses = {", path = "test.dat")
write_lines(x = '<"Anne", 11>', path = "test.dat", append = TRUE)
write_lines(x = "};", path = "test.dat", append = TRUE)

cat(read_file("test.dat"))
```

Load datasets:

```{r}
remotes::install_github("bss-osca/tfa/tfa-package", dependencies = FALSE)

library(tidyverse)
nurses <- read_csv(system.file("extdata/nurses.csv", package = "tfa"))
shifts <- read_csv(system.file("extdata/shifts.csv", package = "tfa"))
```

**Question 2 - Transform all character columns in `nurses` so they start and end with `"`.**

```{r}
head(nurses)

nurses <- nurses %>% 
  mutate(across(where(is.character), ~str_c('"', .x, '"'))) %>% print()
```


**Question 3 - Unite all columns into a new column named `tuple` where each column is separated with `,`. Hint: have a look at the `unite` function. All columns can be selected using `everything()`.**

```{r}
?unite

nurses <- nurses %>% 
  unite(col = "tuple", sep = ", ", remove = FALSE) %>% print()
```


**Question 4 - Add `<` and `>` at the start and end of the `tuple` column.**

```{r}
nurses <- nurses %>% 
  mutate(tuple = str_c("<", tuple, ">")) %>% print()
```


**Question 5 - Extract the `tuple` column and transform it into a string with `collapse = ",\n"`.**

```{r}
tuple <- nurses %>% 
  pull(tuple) %>% 
  str_c(collapse = ",\n") %>% print()
```


**Question 6 - Create a function `write_tuple` that takes nurses as input and write the tuples to a file.**

```{r}
write_tuple <- function(nurses, file) {
  write_lines(x = "nurses = {", path = file, sep = " \n")
  tuples <- nurses %>% 
    pull(tuple) %>% 
    str_c(collapse = ",\n")
  write_lines(x = tuples, path = file, append = TRUE)
  write_lines(x = "};", path = file, append = TRUE)
}

write_tuple(nurses, "nurse_tuple.dat")

cat(read_file("nurse_tuple.dat"))


# Teacher's solution didn't save all the changes we made along the way to nurses
# So he has everything here in the function, instead of just the last changes, which makes sense I guess.
# Would be like this then (clear nurses to vanilla state for this one to work properly):

write_tuple <- function(nurses, file) {
  write_lines(x = "nurses = {", path = file, sep = " \n")
  tuples <- nurses %>% 
    mutate(across(where(is.character), ~str_c('"', .x, '"'))) %>% 
    unite("tuple", sep = ", ", remove = FALSE) %>% 
    mutate(tuple = str_c("<", tuple, ">")) %>% 
    pull(tuple) %>% 
    str_c(collapse = ",\n")
  write_lines(x = tuples, path = file, append = TRUE)
  write_lines(x = "};", path = file, append = TRUE)
}
```


**Question 7 - Modify `write_tuple` so it works if `shifts` are given as input instead of `nurses`.**

```{r}
write_tuple <- function(data, file) {
  name <- deparse(substitute(data))
  write_lines(x = str_c(name, " = {"), path = file, sep = " \n")
  tuples <- shifts %>% 
    mutate(across(where(is.character), ~str_c('"', .x, '"'))) %>% 
    unite("tuple", sep = ", ", remove = FALSE) %>% 
    mutate(tuple = str_c("<", tuple, ">")) %>% 
    pull(tuple) %>% 
    str_c(collapse = ",\n")
  write_lines(x = tuples, path = file, append = TRUE)
  write_lines(x = "};", path = file, append = TRUE)
}

write_tuple(shifts, "shifts_tuple.dat")

cat(read_file("shifts_tuple.dat"))
```


**Question 8 - Modify `write_tuple` with a new input argument `append` which is false by default. If true, then then the file is not overwritten.**

```{r}
write_tuple <- function(data, file, append = FALSE) {
  name <- deparse(substitute(data))
  write_lines(x = str_c(name, " = {"), path = file, sep = " \n", append = append) # Change is here
  tuples <- shifts %>% 
    mutate(across(where(is.character), ~str_c('"', .x, '"'))) %>% 
    unite("tuple", sep = ", ", remove = FALSE) %>% 
    mutate(tuple = str_c("<", tuple, ">")) %>% 
    pull(tuple) %>% 
    str_c(collapse = ",\n")
  write_lines(x = tuples, path = file, append = TRUE)
  write_lines(x = "};", path = file, append = TRUE)
}
```


**Question 9 - Write nurses and shifts to a single data file.**

```{r}
write_tuple(nurses, "nurses_shifts.dat")
write_tuple(shifts, "nurses_shifts.dat", append = TRUE)

cat(read_file("nurses_shifts.dat"))
```



## Colophon 
<!-- Always keep this section for reproducibility -->

This report has been created inside [RStudio](http://www.rstudio.com/ide/) using [R Markdown](https://rmarkdown.rstudio.com/). 

The report was built using:

```{r message = FALSE, warning = FALSE, echo = FALSE}
session <- devtools::session_info()
session$platform
```

Along with these packages:

```{r message = FALSE, warning = FALSE, echo = FALSE}
session$packages %>% 
  select(package, loadedversion, date, source) %>% 
  DT::datatable(rownames = FALSE,
                class = 'cell-border stripe',
                filter = list(position = 'top'),
                options = list(pageLength = 5, 
                           autoWidth = FALSE,
                           bInfo = FALSE,
                           paging = TRUE))
```








